---
title: "Medical Appointment No Shows"
author: "Akash Barnwal"
date: "April 23, 2017"
output: html_document
---


Setting up important libraries
```{r}
library(tidyverse)
library(lubridate)
library(e1071)
library(caret)
library(randomForest)
library(nnet)
library(dplyr)
library(corrplot)
```


Taking data from the UCI machine learning repository. 
```{r}
medical_data <- read.csv("E:/DataScience/Semester 2/Statistical Learning/Mara/Medical_Data (1).csv",header=T, na.strings=c(""))
```


Inference: 
The data contains 3 lakh records of medical appointment with 15 characteristic of each. The target
variable is indicating whether a patient will show up or not show up at the appointment.

The Initial Step in Building Successful Predictive Analytics Solutions

a) Depicting a random sample of 100 records from the file to understand the data
```{r}
medical_data_sample <- medical_data[sample(1:nrow(medical_data), 100, replace=FALSE),]
medical_data_sample
```

b) Checking no of missing values in data
```{r}
sapply(medical_data,function(x) sum(is.na(x)))
```

Inference:

We can see that there are no missing values in the data. The objective of doing this is to remove all the variables which has more than 90% missing values or incase we have few missing 
values then replacing them with mean value or removing them. 

c) Checking for unique values in the data
```{r}
sapply(medical_data, function(x) length(unique(x)))
```

Inference:
With most of the variables showing binary values, appointment registration has 295425 unique records.
Awaiting time has 213 unique records. AppointmentData has 534 records.The objective of doing this to remove all the variables which has one unique values. 

```{r}
summary(medical_data)
```

```{r}
range(medical_data$Age)
```

Inference: 
We can see that age has negative values which we need to remove for better analysis. Removing the age with negative values

```{r}
medical_data <- medical_data[medical_data$Age>0,]
```


Looking into the data we can see some of the values of awaiting time to be negative which needs to be taken as positive values as Time is a positive value.

```{r}
medical_data$AwaitingTime <- abs(medical_data$AwaitingTime)
```


```{r}
summary(medical_data)
```


d) Lets have a look into duplicate values which could be there and can be responsible for deviating the analysis

```{r}
dup_rows <- duplicated(medical_data)
dup_rows_num <- sum(dup_rows)
dup_rows_num
```

Inference: 
There are a total of 322 records which are duplicate values. A possible reason could be two or more appointment
for the same person.

General Data Exploration


Let's see the distribution of "No-Show" and "Show-up" cases
```{r}
status_matrix <- table(medical_data$Status)
status_matrix
```

```{r}
ggplot(medical_data, aes(x=Status, fill=Status)) + geom_bar()
```


Inference: 
A total of 201889 records are show up and 87779 are no show up.


The percentage of people who don't show it is very high:
```{r}
medical_data %>% group_by(Status) %>% summarise(Count = n()) %>% mutate(Percent = Count/sum(Count))
```


Now lets explore the analysis on "AppointmentData":

```{r}

medical_data %>% group_by(ApointmentData) %>% summarise(total_noshow=sum(Status=="No-Show")/n()) %>% ggplot(aes(x=ApointmentData, y=total_noshow)) + geom_point(alpha=0.3) + geom_smooth(method = "lm")
```


Inference: 
With all the data at the center it can be said that the proportion of "No-show" remains constant throughout the time period. 


```{r}


medical_data %>% group_by(Gender) %>% summarise(Count = n()) %>% mutate(Percent = Count/sum(Count))
```
The total percentage of female count being 195400 is high as compared to male count.




```{r}
library(dplyr)
library(tidyr)
show_noshow <- medical_data %>%  group_by(Age, Gender) %>%  count(Status)

show_noshow <- spread(show_noshow, key = Status, value = n)
```

```{r}
library(knitr)
colnames(show_noshow) <- c("Age", "Gender", "Noshow", "Show")
```

Calculating a no show rate
```{r}

library(tidyr)
show_noshow <- show_noshow %>%  mutate(Noshowrate = round(Noshow/(Noshow + Show), digits = 3))
```

```{r}
show_noshow$Gender <- factor(show_noshow$Gender, levels = c("M", "F"))
```


```{r}
show_noshow %>%
  group_by(Age) %>% 
  summarise(Noshowrate = mean(Noshowrate)) %>% 
  ggplot() +
  aes(x = Age, y = Noshowrate) +
  geom_bar(stat = "identity", fill = "gray47") +
  xlab("Age") +
  ylab("No-Show Rate") +
  ggtitle("No-Show Rate Depends on Age") 
```
 
Inference:
These plots both show that the younger a person is, the more likely they will not show up and that there are a couple of outliers in age. The no-show rate of medical appointments is highly dependent on the age of the appointee, with the rate maxed out around 40% of cancellations in ages from 14 to 24 years old. The no-show rate then decreases with a minimum around 20% with 70 year old appointees. After that, the rate increases again, which may be due to smaller sample sizes compared to other ages.


```{r}
library(dplyr)
show_noshow %>% 
  ggplot() +
  aes(x = Age, y = Noshowrate, fill = Gender) +
  geom_bar(stat = "identity") +
  facet_wrap(~Gender) +
  xlab("Age") +
  ylab("No-Show Rate") +
  ggtitle("No-Show Rate Depends on Age, Male / Female") +
  scale_fill_manual(values = c("cadetblue4", "hotpink2")) +
  
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "gray96"),
        axis.title.x=element_blank())
```

According to the graph it can be seen that there is no significant difference between male and female for no show rate.

Lets do a similar analysis for the weeks.
```{r}
 weekday <- medical_data %>% group_by(DayOfTheWeek, Gender) %>% count(Status)

 weekday <- spread(weekday, key = Status, value = n)
 
 colnames(weekday) <- c("DayOfTheWeek", "Gender", "Noshow", "Show")
 
 weekday <- weekday %>% mutate(Noshowrate = round(Noshow / (Noshow + Show), digits = 3))
 
 weekday$Gender <- factor(weekday$Gender, levels = c("M", "F"))
```


```{r}

 weekday %>%
  group_by(DayOfTheWeek) %>% 
  summarise(Noshowrate = mean(Noshowrate)) %>%
  ggplot() +
  aes(x = DayOfTheWeek, y = Noshowrate) +
  geom_bar(stat = "identity", fill = "gray47") +
  scale_x_discrete(limits = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
                   labels = c("Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun")) +
  xlab("Day of Week") +
  ylab("No-Show Rate") +
  ggtitle("No-Show Rate Against Day of Week") +
  
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "gray96"),
        axis.title.x=element_blank())
```



Similar analysis were done in regards of the day of the week. Overall, the no-show rate stays quite similar (~30%) across weekdays. Saturday's no-show rates are higher, however the number of appointments scheduled on Saturdays are also smaller (around 1400 total, compared to 10k+ for all other days), the same goes with Sunday appointments (only 6).


```{r}

  weekday %>% 
  ggplot() +
  aes(x = DayOfTheWeek, y = Noshowrate, fill = Gender) +
  geom_bar(stat = "identity") +
  facet_wrap(~Gender) +
  scale_x_discrete(limits = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
                   labels = c("Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun")) +
  xlab("Day of Week") +
  ylab("No-Show Rate") +
  ggtitle("No-Show Rate Against Day of Week") +
  scale_fill_manual(values = c("cadetblue4", "hotpink2")) +
  
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "gray96"),
        axis.title.x=element_blank())
```


Weekday analysis was then done in regards with the gender of the appointee, however there doesn't seem to be a significant difference. Saturday and Sunday differences also due to a smaller sample size as shown before. 


Renaming each of the variables in the data since we see that there are columns with wrong name such as "Alcoolism" which stands for "Alcoholism". 

```{r}
medical_data<-dplyr::rename(medical_data,
                      AppointmentData=ApointmentData,
                      Alcoholism=Alcoolism,
                      Hypertension=HiperTension, 
                      Handicap=Handcap
                      )
```

```{r}
library(lubridate)
medical_data$AppointmentData <- ymd_hms(medical_data$AppointmentData)
medical_data$AppointmentRegistration<-ymd_hms(medical_data$AppointmentRegistration)
medical_data$DayOfTheWeek<-factor(medical_data$DayOfTheWeek,levels=c('Sunday','Monday','Tuesday','Wednesday','Thursday','Friday','Saturday'))
medical_data$Status<-factor(make.names(medical_data$Status))
medical_data$AwaitingTime<-abs(medical_data$AwaitingTime)
medical_data$Gender<-as.factor(medical_data$Gender)

```

```{r}
g.smokes <- ggplot(medical_data, aes(x=Smokes,fill=Status))+geom_bar(position='fill')+ggtitle('Smokes')+ylab('Proportion')
g.smokes
```

```{r}

g.Tuberculosis<-ggplot(medical_data,aes(x=Tuberculosis,fill=Status))+geom_bar(position='fill')+ylab('Proportion')+ggtitle('Tuberculosis')
g.Tuberculosis
```

```{r}
g.Scholarship<-ggplot(medical_data, aes(x=Scholarship,fill=Status))+geom_bar(position='fill')+ggtitle('Scholarship')+ylab('Proportion')
g.Scholarship
```

```{r}
g.Handicap<-ggplot(medical_data, aes(x=Handicap,fill=Status))+geom_bar(position='fill')+ggtitle('Handicap')+ylab('Proportion')
g.Handicap
```

```{r}
g.Diabetes<-ggplot(medical_data, aes(x=Diabetes,fill=Status))+geom_bar(position='fill')+ggtitle('Diabetes')+ylab('Proportion')
g.Diabetes
```

```{r}
g.Alcoholism<-ggplot(medical_data, aes(x=Alcoholism,fill=Status))+geom_bar(position='fill')+ggtitle('Alchoholism')+ylab('Proportion')
g.Alcoholism
```

```{r}
g.Hypertension<-ggplot(medical_data, aes(x=Hypertension,fill=Status))+geom_bar(position='fill')+ggtitle('Hypertension')+ylab('Proportion')
g.Hypertension
```

```{r}
g.Sms<-ggplot(medical_data, aes(x=Sms_Reminder,fill=Status))+geom_bar(position='fill')+ggtitle('Number of SMS Reminders')+ylab('Proportion')
g.Sms
```

```{r}
# library(gridExtra)
# grid.arrange(g.smokes,g.Tuberculosis,g.Scholarship,g.Handicap,g.Diabetes,g.Alcoholism,g.Hypertension,g.Sms,ncol=2)
```

Lets check the show up and No show up based on weekdays
```{r}
ggplot(medical_data,aes(x=DayOfTheWeek,fill=Status))+geom_bar(position='fill')+ylab('proportion')

```


From the given graph we can say that the number of show-up were higher at Sunday. That makes sense since its a
holiday at every place.The no. of show-up were less for Saturday as compared to other day of the weeks. There could be various reasons for this.

Lets have a look into number of datas for each of the week.
```{r}
table(medical_data$DayOfTheWeek)
```
We dont have much of records for sunday and staurday so a lot of inference about the showup and no showup doesnt make sense for these two days in comparison with other day of the weeks. 


Lets see the show-up and no show-up based on awaiting time
```{r}
ggplot(medical_data,aes(x=AwaitingTime,fill=Status))+geom_histogram(position='fill',binwidth=10)+ylab('proportion')
```

```{r}


   medical_data %>% group_by(Age) %>% summarise(rate=sum(Status=='No.Show')/n()) %>%   
   ggplot(aes(x=Age,y=rate,fill=Age)) + geom_bar(stat='identity')+
  scale_fill_gradient(low="red",high="blue")
```


```{r}
cor.medical_data<-cor(medical_data[sapply(medical_data,is.numeric)])

library(corrplot)
corrplot(cor.medical_data,method='number')
```

Partitioning the data into test and training set. Taking 70 % of the data as train data and 30% as test data
```{r}
library(caret) 
set.seed(1234)
split_data <- createDataPartition(medical_data$Status, p = 0.7, list = FALSE)
train_data <- medical_data[split_data,]
test_data <-  medical_data[-split_data,]
```

```{r}


fitControl <- trainControl(method = "cv",
                           number = 5,
                           #savePredictions="final",
                           summaryFunction = twoClassSummary,
                           classProbs = TRUE
)
```



As we can see that number of "No.Show" is less than "Show" hence we will upscale the  No.Show class


```{r}
train_data <- upSample(train_data[, setdiff(names(medical_data), 'Status')], train_data$Status, yname="Status")
table(train_data$Status)
```


```{r}
fit_rpart <- train(Status~., 
                   train_data,
                   method = "rpart",
                   metric = "ROC",
                   trControl = fitControl)

pred_rpart <- predict(fit_rpart, test_data)
confusionMatrix(pred_rpart, test_data$Status)
```



```{r}
library(caTools)
pred_rpart2 <- predict(fit_rpart, test_data, type="prob")
colAUC(pred_rpart2, test_data$Status, plotROC=TRUE)
```




Lets make a decision tree and see which are the important variables to validate show up and no-show up. 
```{r}
library(rpart.plot)
rpart.plot(fit_rpart$finalModel, type = 2, fallen.leaves = F, cex = 1, extra = 2)
```

We have tried rpart with other feature variable combinations but there are not great difference.


Looking at the model we can state that older people with age greater than 46 tend to show up always. Incase
of younger people, they show up if the appointment is close to 8.5 days or less. 


Lets run a logistic regression on the data
```{r}
fit_glm <- train(Status~.-AppointmentRegistration-AppointmentData-Handicap-Tuberculosis, 
                 train_data,
                 method = "glm",
                 metric = "ROC",
                 preProc = c("center", "scale"),
                 trControl = fitControl)

summary(fit_glm)

```
Inference: 

a) We can see that the variables Gender, Diabetes are not statistically significant. 
b) As for the statistically significant variables, Alcoholism and Age has the lowest p-value 
   suggesting a strong association of the Alcholism and Age of the  
   patient with the probability of not showing up.
   

```{r}
pred_glm <- predict(fit_glm, test_data)
confusionMatrix(pred_glm, test_data$Status)
```


```{r}
summary(fit_glm$finalModel)
pred_glm2 <- predict(fit_glm, test_data, type="prob")
colAUC(pred_glm2, test_data$Status, plotROC=TRUE)
```
This model leds an auc greater but near to the rpart model one.

This logist regression model says that these variables are significant and increase the risk of "No Show":


```{r}
summary(fit_glm)$coef %>% 
as.data.frame() %>% 
cbind(feature=rownames(summary(fit_glm)$coef)) %>% 
filter (.[[4]] <= 0.05) %>% 
arrange(desc(abs(Estimate)), desc(.[[4]]))
```

The important variables which affect the probability of "No Show" are :

a) Age: Younger people affect the no show as compared to older people
b) Awaiting time: The greater the awaiting time, the chances of no show up increases
c) Scholarship: Wherever there is an existent of Scholarship, the probability of no-show gets affected

Age and AwaitingTime coincide with the classification tree model results. It is surprising that Gender is not significative enough to be considered as a important.

```{r}
medical_data$AwaitingTime <- abs(medical_data$AwaitingTime)
medical_data$Sex <-ifelse(medical_data$Gender =="F", 1, 0)
is.numeric(medical_data$Sex)
```

```{r}

medical_data$Status<- as.factor(medical_data$Status)
medical_data$Visit<- ifelse(medical_data$Status == "Show-Up", 1, 0)
is.numeric(medical_data$Visit)
```



